Global patterns of forest autotrophic carbon fluxes

### 2020-05-04 manuscript/manuscript draft.Rmd
Update manuscript draft.Rmd
worked on discussion
#69

Length of the growing season was estimated to the nearest month, where growing season months were defined as months with mean minimum temperature > 0.5$^\circ$C. We experimented with a definion of growing season months including a moisture index, defined as (MAT - PET)/PET, > -0.95 (Kerkhoff et al. 2005; see also Michaletz et al. 2014). However, we found that including a moisture index had **no** effect on the estimates of growing season length, and so chose to exclude it.  (**Becky, was it really no effect? or minimal?**) Monthly data for PET, precipitation, and temperature from the CRU dataset v 4.03 [@harris_updated_2014], and solar radiation from WorldClim2 [@fick_worldclim_2017] were used to calculate mean monthly PET, precipitation, temperature and solar radiation during the growing season. Total growing season precipitation and solar radiation were also calculated.

@@

**This paragraph needs work--including re-assessment of results, interpretation, writing**
We show that, with the exception of some differences in MAT-MAP interactions (Fig. 3), FACF are broadly consistent in their responses to climate drivers on the global scale, with no pronounced trends in C allocation among the variable pairs tested (Figs. 2, S3). Although variation in allocation has been observed along gradients of elevation [@moser_elevation_2011] and water availability [@newman_above-_2006]--along with non-climatic axes of stand age [@litton_carbon_2007], nutrient availability [@litton_carbon_2007; @gill_belowground_2016], and forest structure [@taylor_greater_2019]--little variation in relation to climate is apparent at the global scale. Our conclusion, then, is that hypothesized gradients in allocation along global climate gradients cannot currently be supported for mature forests, although data quantity and standardization is currently insufficent to rule out the possibility that such trends exist. There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there are no signficant trends in allocation of $GPP$ to $NPP$ or of $NPP$ to $ANPP_{woody-stem}$ across gradients of latitude or climate (Fig. S3). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

**there seems to be a difference between GPP and other fluxes at higher precips (Figs. 3, 4b), but the only allocaiton ratio we consider is GPP:NPP. It would be interesting too consider some others. Perhaps precip-->soil fertility-->CUE (see Doughty et al. 2018). I need to come back to this.**

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$ - latitude $R^2\ge$ 48%) than in subsidiary fluxes (latitude $R^2$ <27% for $BNPP_{fine.root}$, $R_{auto-root}$,$ANPP_{woody-stem}$) (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (**bbl: cite e.g. Huntzinger?**). However, methodological variation and uncertainty affect all of fluxes considered here--not necessarily any less than the aforementioned, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more dicrect climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, and secondarily by moisture availability (Table 1, *H2-H3*; Figs. 3-4). Because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means, it is not appropriate to attempt to identify any one mean annual climate variable as a mechanistic driver of FACF. However, it remains informative to consider these relationships. We find that temperature-related climate variables ($MAT$, temperature seasonality, ...**LIST**) explain the highest proportion of variability in FACF, and among these, $MAT$ is generally the best predictor--perhaps because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables are extracted from global gridded data products (Table S1). The effects of temperature are modified by moisture availability, with reduced FACF under hot and dry conditions (*i.e.*, high PET, high deficit; Fig. 4c-d) and sometimes under very high precipitaiton (Figs. 3, 4b). Negative effects of very high precipitation on FACF have been observed previously (**REFS**) and are attributable to nutrient and light limitations (**REFS**). Thus, although temperature and water availability jointly and interactively drive global-scale patterns of FACF.

FACF are negatively correlated with temperature seasonality (Table 1, *H4*; Fig. 4e), and is minimal during cold- or dry- dormant seasons. To account for this, a number of analyses seeking to characterize global-scale effects of climate on productivity have examined the relationship of C flux per month of the growing season with growing season climatic conditions (Table 1, *H5*; **REFS**). We found that the sort of simple metric needed to define growing season at a global scale was uncertain for temperature and problematic for moisture (**WORK ON THIS**). A temperature-defined growing season length had stong positive correlation with FACF (Fig. 4f), but explained less variation than $MAT$. Dividing FACFs by growing season length to yield FACF per growing season month removed the majority of climate-related variation, supporting the idea that the latitudinal gradient in FACF is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosyntheiss or respiration by high-latitude forests (*[ Enquist et al. 2007 GCB- but check]*). However, there remained a number of significant correlations with growing season climatic conditions, suggesting that climatic conditions remain influential within the growing season. We conclude that while correcting for growing season length takes analyses a step closer to mechanistic linkage of instantaneous C flux rates to environmental conditions, it remains very crude relative to the the timescales on which climate affects plant metabolism and does not advance statistical predictive power. Rather, mechanistic accounting for climatic effects on global FACF patterns requires models representing physiologically meaningful timescales (*e.g.*, **refs**).

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ [@li_mapping_2019] and $ANPP_{woody stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. However, increasing C flux rates are not universal [@rutishauser_testing_2020] (**MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of cliamte, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

>>>>

Length of the growing season was estimated to the nearest month, where growing season months were defined as months with mean minimum temperature > 0.5$^\circ$C. We experimented with a definion of growing season months including a moisture index, defined as (MAT - PET)/PET, > -0.95 (Kerkhoff et al. 2005; see also Michaletz et al. 2014-**put in these refs**). However, we found that including a moisture index had **no** effect on the estimates of growing season length, and so chose to exclude it.  (**Becky, was it really no effect? or minimal?**) Monthly data for PET, precipitation, and temperature from the CRU dataset v 4.03 [@harris_updated_2014], and solar radiation from WorldClim2 [@fick_worldclim_2017] were used to calculate mean monthly PET, precipitation, temperature and solar radiation during the growing season. Total growing season precipitation and solar radiation were also calculated.

@@

**See [issue #70](https://github.com/forc-db/Global_Productivity/issues/70), and potentially rework paragraph. **
We show that, with the exception of some differences in MAT-MAP interactions (Fig. 3), FACF are broadly consistent in their responses to climate drivers on the global scale, with no pronounced trends in C allocation among the variable pairs tested (Figs. 2, S3). Although variation in allocation has been observed along gradients of elevation [@moser_elevation_2011] and water availability [@newman_above-_2006]--along with non-climatic axes of stand age [@litton_carbon_2007], nutrient availability [@litton_carbon_2007; @gill_belowground_2016], and forest structure [@taylor_greater_2019]--little variation in relation to climate is apparent at the global scale. Our conclusion, then, is that hypothesized gradients in allocation along global climate gradients cannot currently be supported for mature forests, although data quantity and standardization is currently insufficent to rule out the possibility that such trends exist. There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there are no signficant trends in allocation of $GPP$ to $NPP$ or of $NPP$ to $ANPP_{woody-stem}$ across gradients of latitude or climate (Fig. S3). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$) than in subsidiary fluxes ($BNPP_{fine.root}$, $R_{auto-root}$, $ANPP_{woody-stem}$) (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (*Clark et al. 2001 Ecological Applications*). However, methodological variation and uncertainty affect all of fluxes considered here--not necessarily any less than the aforementioned, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

**This paragraph needs work, but I'm not currently sure how to fix it. It reads more as results than discussion, and doesn't say anything really profound. Depending on what happens with [issue #70](https://github.com/forc-db/Global_Productivity/issues/70), we might somehow rework/rearrange this paragraph and the previous two. **
The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, and secondarily by moisture availability (Table 1, *H2-H3*; Figs. 3-4). Because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means, it is not appropriate to attempt to identify any one mean annual climate variable as a mechanistic driver of FACF. However, it remains informative to consider these relationships. We find that temperature-related climate variables ($MAT$, temperature seasonality, ...**LIST**) explain the highest proportion of variability in FACF, and among these, $MAT$ is generally the best predictor--perhaps because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables are extracted from global gridded data products (Table S1). The effects of temperature are modified by moisture availability, with reduced FACF under hot and dry conditions (*i.e.*, high PET, high deficit; Fig. 4c-d) and sometimes under very high precipitaiton (Figs. 3, 4b). Negative effects of very high precipitation on FACF have been observed previously (**REFS**) and are attributable to nutrient and light limitations (**REFS**). Thus, although temperature and water availability jointly and interactively drive global-scale patterns of FACF....

FACF are negatively correlated with temperature seasonality (Table 1, *H4*; Fig. 4e), and is minimal during cold- or dry- dormant seasons. To account for this, a number of analyses seeking to characterize global-scale effects of climate on productivity have examined the relationship of C flux per month of the growing season with growing season climatic conditions (Table 1, *H5*; **REFS**). We found that the sort of simple metric needed to define growing season at a global scale [@kerkhoff_plant_2005;@michaletz_convergence_2014] was rough with respect to temperature and problematic with respect to moisture because it doesn't capture temporal lags between precipitation and plant water use caused by storage in soil or snow. (**probably better to get into this in the methoods**). A temperature-defined growing season length had stong positive correlation with FACF (Fig. 4f), but explained less variation than $MAT$. Dividing FACFs by growing season length to yield FACF per growing season month removed the majority of climate-related variation, supporting the idea that the latitudinal gradient in FACF is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosyntheiss or respiration by high-latitude forests (*[ Enquist et al. 2007 GCB- but check]*). However, there remained a number of significant correlations with growing season climatic conditions, suggesting that climatic conditions remain influential within the growing season. We conclude that while correcting for growing season length takes analyses a step closer to mechanistic linkage of instantaneous C flux rates to environmental conditions, it remains very crude relative to the the timescales on which climate affects plant metabolism and does not advance statistical predictive power. Rather, mechanistic accounting for climatic effects on global FACF patterns requires models representing physiologically meaningful timescales (*e.g.*, **refs**).

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ [@li_mapping_2019] and $ANPP_{woody stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. However, increasing C flux rates are not universal [@rutishauser_testing_2020] (**MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.
